---
title: "When to use the residual RFLVM (and when not)"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

**Author:** Jun Inamo (juninamo@keio.jp)

The residual random-feature latent variable model (residual RFLVM:
`rff_programs()`, `fit_spatial_rff()` with an offset, `rff_program_test()`,
`rff_transfer_test()`, `rff_lengthscale_profile()`) is **experimental**. This
page says what it is good for, based on evaluations on real synovial Xenium
5K data, in vitro Xenium data sets and Visium HD, and where simpler tools are
better. The code chunks need real data and are not run.

## In one paragraph

The residual RFLVM is a **discovery-and-transfer tool**. Use it to find
multi-gene spatial programs beyond the structure you already know (cell
types, domains, distances), without pre-specifying genes or structures, and
then to check that a program is real by a **calibrated transfer** to held-out
patients or an independent cohort. It is not the best tool for a question you
can already state (distance to a structure, a known gene set), it does not
measure signalling reach as a distance at cellular resolution, and on its own
it has no power for clinical prediction in small cohorts.

## Decision table

| Your question | Use | Why |
|---|---|---|
| Which genes change with distance to a known structure (vessel, lining, TLS)? | Per-gene GLM on distance (e.g. quasi-Poisson with window or patient effects), C-SIDE | Supervised methods recovered a known perivascular gradient directly and matched or beat the RFLVM in spike-ins (gene AUC 0.82-0.85 vs 0.70-0.73) |
| Do two cell types or gene sets co-localize, at which distance, and does it differ between groups? | `nhood_enrichment()`, `cooccur_local_oe()`, `pcf_cross()`, `pcf_matrix()`, `compare_groups()` | Calibrated O/E effect sizes and patient-level tests; no model needed |
| Is a known gene set spatially patterned? | Score the set and use `pcf_cross()`, or `rff_transfer_test()` with the set as fixed loadings | A pre-specified hypothesis needs no discovery step |
| Are there multi-gene spatial programs beyond known structure, without naming genes or structures? | **Residual RFLVM**: `rff_programs()` with a rich offset, control calibration, `rff_report()` | Unsupervised; separates compartments that gene scans mix |
| Does a program found in one cohort recur in held-out patients, another panel or another platform? | **`rff_transfer_test()`** with fixed loadings | Permuted-loading null keeps each gene's own spatial structure |
| How far does a response reach? | Producer-conditioned regression when producers are known; `rff_lengthscale_profile()` + `rff_reach()` only as a **ranking** | Profiled reach ranks programs at 16-32 um bins; it is not a measured distance at 8 um bins |
| Does a spatial program predict response or outcome? | Not the RFLVM at n < ~50 patients; project loadings onto large bulk cohorts instead | Existing features were at chance with 15 patients; AUC >= 0.86 would be needed for 80% power |

## Recommended workflow

### 1. Put everything you know into the offset

The residual is only as interesting as the offset is rich. With lineage
labels alone, discovered programs were mostly cell-type or sub-lineage
residuals. Use fine labels (sub-clusters) where you have them, a
neighbour / spill-over term, the nuclear share, and structure covariates.

```{r offset}
b  <- bin_transcripts(tx, bin_size = 8)
E  <- rff_expected_offset(b, tx, type_col = "fine_type", neighbour_bandwidth = 10)
off <- rff_offset(b, covariates = data.frame(inside_tls, border_dist, nuclear_share),
                  base = E,
                  gene_covariates = list(spill = attr(E, "neighbour") - E))
```

Leave out the covariate you want to discover (e.g. do not put vessel distance
into the offset if the question is whether a perivascular program exists).

### 2. Build a negative-control reference on the same tissue

On real tissue the plain gene-shift test (`null = "shift"`) is calibrated
against its own null but calls almost every window, because real tissue
always carries shared residual structure. Choose regions where the program of
interest is not expected (tissue far from the structure, a homogeneous
monolayer far from the inducing cells, untreated wells), fit them with the
same settings, and use at least 19 of them.

```{r controls}
ctrl_fits <- lapply(seq_along(ctrl_b), function(i)
  rff_programs(ctrl_b[[i]], offset = ctrl_off[[i]], null = "none")$fit)
ref <- rff_control_reference(ctrl_fits, ctrl_b, highpass = 32)
```

### 3. Discover programs with the calibrated test

```{r discover}
res <- rff_programs(b, offset = off, lengthscale = 8, highpass = 32,
                    crossfit = TRUE, control = ref,
                    report = "programs.html")
res
```

For several windows or patients, `fit_spatial_rff_joint()` and
`rff_program_test_joint(crossfit = TRUE, control = ref)` give shared
loadings. Split patients into discovery and validation sets **before** this
step.

### 4. Inspect every program before interpreting it

`rff_report()` (or `report =` above) shows the top genes, maps, which cell
labels own the excess transcripts, and the correlation with known programs
(`known_programs =`). Ask for each program:

* Is it a cell-type or sub-lineage residual (top genes are markers of one
  lineage; excess transcripts sit in neighbouring lineages)?
* Is it technical (loadings track the genes' nuclear fraction, total counts,
  or spill-over from a neighbouring compartment)?
* Is it a known program (does it vanish when the known gene set is added to
  the offset)?

### 5. Validate by transfer to held-out data

```{r transfer}
L <- res$fit$L[, c("factor2", "factor4")]          # programs of interest
tt_target  <- rff_transfer_test(L, val_fits,  val_b,  n_perm = 999)
tt_control <- rff_transfer_test(L, ctrl_val_fits, ctrl_val_b, n_perm = 999)
```

A program is supported when it transfers to held-out patients or another
cohort **and** is more active in target than in control regions. Transfer to
every region of a platform is not evidence of specificity.

### 6. Reach: ranks and identifiable ranges only

```{r reach}
b16 <- bin_transcripts(tx, bin_size = 16)
fit16 <- fit_spatial_rff(b16, n_factors = 6, offset = off16, basis = "grid",
                         lengthscales = "profile", ls_grid = 5 * 2^(1:7),
                         profile = list(init = 32))
fit16$lengthscale_profile$summary[, c("factor", "reach", "reach_lower",
                                      "reach_upper", "identifiable")]
```

Profile at two or three bin sizes (16-32 um), report the order of reaches,
and treat a value as a distance only when `identifiable` is `TRUE`.

## Pitfalls

* **Uncalibrated tests.** Without `control =`, "significant" means "some
  shared residual structure exists", which is true almost everywhere.
* **Significance is not novelty.** Recurrent programs that replicate across
  folds, cohorts and offsets can still be cell-type residuals.
* **Reach at fine bins.** At 8 um bins the profile returns 25-50 um for any
  true reach, because single cells dominate the correlation.
* **Amplitude is not level.** A program's amplitude measures patterning
  within a window; a simple per-cell readout of its genes can be the better
  group statistic.
* **Batch and panel.** Program amplitudes move with panel and batch; compare
  cohorts only with a design that separates them.
* **Small cohorts.** Per-window effects with many windows per patient are not
  per-patient effects; test at the patient level.

## Validation evidence (summary)

All numbers from real-data evaluations in 2026 (RA and JIA synovium, Xenium
5K; in vitro Xenium; Visium HD); no individual-level data are shown.

| Question | Result |
|---|---|
| Plain gene-shift test on real tissue | Gene-shifted data called in 0/46 windows (null is valid), but real TLS windows were called in 95% and negative-control tiles in 42% |
| Control-calibrated, cross-fitted test | Negative-control tiles called 0/24 (Xenium 5K), 1/24 (coculture), 0/24 (Visium HD), 1/24 (organoid); planted 6-gene programs found in 33 / 58% (log-amplitude 0.75 / 1) on real tiles; power depends on window size |
| Transfer test (score statistic) | Simulation: 20/20 true transfers, 0/20 without the program, 0/20 with another program |
| Length scales | Learned length scales stay at their initial values; profiled reach ranks spike-ins at 16 um bins (Spearman 0.95, co-localised programs ordered 6/6) but is compressed (25-95 um); within 1.5x of the truth only for 40-160 um at 32 um bins in 3.2 mm windows; unresolved at <= 20 um |
| Perivascular program (known gradient) | A per-gene distance GLM recovered the published Notch / TGF-beta gradient directly; in a spike-in benchmark no method detected the weak spike, and supervised methods matched or beat the RFLVM on gene AUC |
| Perivascular program (discovery + transfer) | Without specifying genes or vessels, the RFLVM found a fibroblast-owned NOTCH3 / HEY1 / NOTCH4 program 0-30 um from arterioles, distinct from the vessel wall and a distal matrix program that a gene scan mixed (gene-scan ranking rho 0.72 with the vessel-wall program, 0.10 with the fibroblast program); it was perivascular-specific in held-out patients (AUC 0.64 vs control windows) and transferred with fixed loadings to an independent cohort on a 50-gene panel in perivascular (z 4.7, p = 0.001) but not control windows (p = 0.10) |
| TLS programs | Eight programs replicated across patient folds, diseases, TLS definitions and offsets, but all were cell-type / sub-lineage residuals, a technical nuclear-RNA axis, or the known IFN-gamma program; adding spill-over and nuclear-share terms did not remove them, and they tracked known sub-lineage states (e.g. tissue-resident macrophages, plasma-cell maturity) |
| Genetic heritability | No TLS program carried RA or JIA heritability beyond cell identity (S-LDSC, smallest q 0.83); cell-type annotations did |
| Clinical prediction | 15 patients: existing features at chance (leave-pair-out AUC 0.20-0.61); AUC >= 0.86 needed for 80% power |
| Multinomial likelihood | No reduction of false calls compared with the negative binomial model |

See the [algorithm reference](algorithms.html) for the model, and
`?rff_programs`, `?rff_program_test`, `?rff_transfer_test` and
`?rff_lengthscale_profile` for details.
